\sysfull

### 2018-04-16 paper/paging.tex

Paging moves data between non-volatile and volatile memory. A page has a home
location in non-volatile memory which we refer to as the page's {\tt store}.
As a task accesses pages, they are swapped from their home location
in the direct\hyp{}mapped non-volatile memory, called
\texttt{\underline{store}} (underlined locations are non-volatile) 
into the fully\hyp{}associative \texttt{working} page buffer in volatile memory. The
\texttt{working} buffer is allocated to take up the volatile memory that
remains after space for stack is allocated.
%
\sys uses two-phase commit to write dirty pages from the \texttt{working} buffer
back to their home location in \texttt{\underline{store}} atomically,
ensuring data consistency even with power failures.
%
When any dirty page is swapped out from the \texttt{working} buffer due to
eviction or task transition from last coalesced task, the first phase writes
the page to a direct\hyp{}mapped non-volatile \texttt{\underline{shadow}}
buffer, and the second phase writes it from
\texttt{\underline{shadow}} to \texttt{\underline{store}}.
%
Unlike prior work~\cite{chain,alpaca} that moves scattered variables one at a
time between volatile and non-volatile memory, \sys's page-based
design moves contiguous blocks of data.
%
\sys takes full advantage of hardware\hyp{}accelerated block copies using
Direct Memory Access (DMA) hardware in common microcontrollers --- e.g. 5x
fewer cycles on MSP430FRxxxx.

@@

A \sys task must access protected non-volatile variables through a restricted memory
interface. The interface includes \texttt{RVAR(var)}, which reads the value of
variable {\tt var} and \texttt{WVAR} \texttt{(var,val)}, which assigns
value {\tt val} to variable {\tt var}. The implementation of {\tt
RVAR} is shown in Algorithm~\ref{algo:rwar} and {\tt WVAR} implementation is
similar except for an assignment instead of the return.
%
{\tt RVAR} and {\tt WVAR} operations translate a variable's physical address in non-volatile memory into a \emph{virtual address} in the page buffer in volatile memory.  In \sys, a virtual address is composed of a \emph{page tag} that identifies the page and a \emph{page offset} that identifies a byte.

@@

\begin{algorithm}[t]
	\caption{\texttt{RVAR}(variable $v$)}
	\label{algo:rwar}
	\scriptsize
	%\small
	\begin{algorithmic}[1]
		\State $t \leftarrow \Call{GetTag}{v}$ 
        \State $i \gets \{ j\ |\ \Call{GetTag}{\texttt{working}[j]} = t \}$ \Comment{Look in the page buffer}
		\If { $i = \emptyset$ } \Comment{Is the variable in a resident page?}
		\State	$i \gets \Call{PageFault}{t}$ \Comment{Page in from non-volatile memory}
		\EndIf
		\State $o \leftarrow \Call{GetOffset}{v}$ 		
		\State \Return \texttt{working}[$i$][$o$]  \Comment{Return directly from page buffer}
	\end{algorithmic}
\end{algorithm}

After address translation, \sys attempts to access the protected variable's
location in the volatile paging buffer. \sys keeps track of the page tags for
the pages currently resident in the paging buffer. When a task accesses a
variable, it compares the variable's page tag to tags of the pages in {\tt
working} (Line 2).
%
%If the accessed variable's page is resident in the page buffer,
%then the tag of one of the pages in this buffer is equal to the accessed
%variable's page tag.
%
If the accessed variable's page tag is not found in the page buffer, the
operation incurs a {\em page fault} (Line 4). At a page fault, \sys must swap
out one of the pages that is resident in the page buffer (to the non-volatile
commit buffer, \texttt{\underline{shadow}}) and swap in the variable's page.
%
The byte is accessed in the page buffer at the index of the resident page and
the variable's page offset (Lines 5-6).

@@

\begin{algorithm}
	\caption{\texttt{PageFault}(tag $t$)}
	\label{algo:pagefault}
	\scriptsize
	%\small
	\begin{algorithmic}[1]
        \State \LeftComment{$i_\text{victim}$: volatile index to next free slot or page to be evicted, set to 0 at boot}
        \State \LeftComment{\texttt{\underline{numShadowed}}: non-volatile, reset to 0 on boot unless in phase 2 commit}
        \State $t_\text{victim} \gets \Call{Tag}{\texttt{working}[i_\text{victim}]}$
		\If {\Call{IsDirty}{working[$i_\text{victim}$]} }	\Comment{Are we evicting a dirty page?}
		\State $\texttt{\underline{shadow}}[t_\text{victim}] \xleftarrow{\text{DMA}} \texttt{working}[i_\text{victim}]$
            \Comment{Swap out the page}
        \If {$t_\text{victim} \not\in \texttt{\underline{shadowList}}$} \Comment{Keep track as list for quick traversal}
            \State $\texttt{\underline{shadowList}}[\texttt{numShadowed++}] \gets t_\text{victim}$
        \EndIf
		\EndIf
		\If {\Call{IsValid}{\underline{shadow}[$t$]}} \Comment{Had the page been swapped out earlier?}
		\State $\texttt{working}[i_\text{victim}] \xleftarrow{\text{DMA}} \texttt{\underline{shadow}}[t]$
            \Comment{Get page from commit buffer}
		\Else
		\State $\texttt{working}[i_\text{victim}] \xleftarrow{\text{DMA}} \texttt{\underline{store}}[t]$
            \Comment{Get page from backing store}
		\EndIf 
        \State $i_\text{victim} \gets (i_\text{victim} + 1) \mod |\texttt{pageBuf}|$ \Comment{FIFO eviction policy}
	\end{algorithmic}
\end{algorithm}

A page fault on a memory access requires \sys to swap out one of the pages in
the page buffer, i.e. \emph{victim page}, preserving updates made to that page,
and to copy the accessed page of protected data into the page buffer.
Algorithm~\ref{algo:pagefault} shows how \sys handles a page fault. A victim
page in the \texttt{working} buffer is selected for eviction according to
\emph{first-in-first-out} replacement policy (Lines 2 and 12).  If the page
being swapped out is not dirty and the page being accessed had not been dirtied
and swapped out to the \texttt{shadow} buffer since last boot, then
handling a page fault consists of copying the accessed page from
\texttt{\underline{store}} into the \texttt{working} buffer (Line 11).
Otherwise, the page fault handler must take the following extra steps.

If a task modified any byte in the victim page
(i.e.,  using \texttt{WVAR}), then the page is dirty. Before swapping in the
accessed page, \sys must first preserve the updated values in the dirty page so
that they can be committed back to their locations in memory
when the dynamic task ends (but not earlier). To preserve dirty values, \sys
copies the victim page to the non-volatile \texttt{\underline{shadow}}
buffer (Line 5) and records the page's tag in a non-volatile list (Lines 6-7)
to support iteration over shadow pages having traversing all slots in the
buffer, necessary in the second phase of the commit.
%
%Dirty pages remain in {\tt \underline{shadow}} buffer until the
%dynamic task completes, at which point \sys commits the modified
%pages from {\tt \underline{shadow}} back to \texttt{\underline{store}}.
%
If the accessed page was previously modified and
swapped out since the last boot (Line 8), the most recent contents of the page is in {\tt
\underline{shadow}}, not \texttt{\underline{store}} (Line 9).

@@

When the last task in a coalesced group completes, \sys must commit \emph{all}
dirty pages in the \texttt{working} buffer and the swapped out pages in
\texttt{\underline{shadow}} buffer back to their original locations in
\texttt{\underline{store}}. If a task is allowed to (re-)execute after only
\emph{some} pages have been copied to \texttt{\underline{store}}, that task
might see protected variables in an inconsistent state.

To make the commit atomic, \sys copies the pages in two phases, implemented by
Algorithm~\ref{algo:commit}.
%
The first phase copies dirty pages in \texttt{working}
buffer to the non-volatile \texttt{\underline{shadow}} buffer (Line 4). 
The second phase, copies pages from \texttt{\underline{shadow}} to \texttt{store} (Line 12).
If power fails during the first phase, the whole
commit is aborted, and the dynamic coalesced task restarts from its first
static task. If power fails in the second phase, the commit will be safely
resumed on next boot.
The non-volatile metadata for the second phase is the \texttt{\underline{committing}}
bit, which is set before the first page is copied (Line 9) and cleared after the last
page is copied (Line 14), and \texttt{\underline{numShadowed}} counter, which
indexes into the \texttt{\underline{shadowList}} (Line 11), is
decremented after each page is copied (Line 13), and cleared once the phase
completes (Line 18).

\sys non-volatile \texttt{\underline{shadowList/numShadowed}} implementation is
a more efficient alternative to a simpler implementation that keeps a valid bit
in each entry of \texttt{\underline{shadow}} and, on every boot into second
phase, traverses all entries, copying the pages with a set bit, and
clearing the bit.
%
\sys assumes that the write to a non-volatile word in memory is atomic.

\begin{algorithm}[t]
	\caption{Two-phase commit}
	\label{algo:commit}
	\scriptsize
	%\small
	\begin{algorithmic}[1]
        \Procedure{CommitPhase1}{} \Comment{Invoked from \texttt{next\_task} on last coalesced task}
            \For {$i \in 0..|\texttt{working}|-1$}
                \State $t \gets \Call{Tag}{\texttt{working}[i]}$
                \State $\texttt{\underline{shadow}}[\Call{Tag}{\texttt{working}[i]}] \xleftarrow{\text{DMA}} \texttt{working}[i]$
                \State $\texttt{\underline{shadowList}}[\texttt{\underline{numShadowed}}] \gets t$
                \State $\texttt{\underline{numShadowed}} \gets \texttt{\underline{numShadowed}} + 1$
            \EndFor
            \State \Call{CommitPhase2}{}
        \EndProcedure
        \Procedure{CommitPhase2}{}
            \State $\texttt{\underline{committing}} \gets \textbf{true}$
            \While{\texttt{\underline{numShadowed}} > 0}
                \State $t \gets \texttt{\underline{shadowList}}[\texttt{\underline{numShadowed}} - 1]$
                \State $\texttt{\underline{store}}[t] \xleftarrow{\text{DMA}} \texttt{\underline{shadow}}[t]$
                \State $\texttt{\underline{shadow}}[t] \gets \textbf{nil}$ \Comment{Clear entry for page fault handler}
                \State $\texttt{\underline{numShadowed}} \gets \texttt{\underline{numShadowed}} - 1 $
            \EndWhile
            \State $\texttt{\underline{committing}} \gets \textbf{false}$
        \EndProcedure
        \Procedure{OnBoot}{} \Comment{Invoked on every boot, before transitioning to a task}
            \If { \texttt{\underline{committing}} } \Call{CommitPhase2}{}
            \EndIf 
            \State $\texttt{\underline{numShadowed}} \gets 0$, $i_\text{victim} \gets 0$
        \EndProcedure
	\end{algorithmic}
\end{algorithm}

@@

\textbf{Memory Consistency.} \sys's paging mechanism ensures that a task only ever executes using consistent
protected data. During task execution, modifications to protected data do not
affect their original memory locations, because a task reads and writes the
volatile \texttt{working} buffer only, and modified pages are kept in
\texttt{\underline{shadow}} until commit. A power failure erases the contents
of the page buffer, preventing a re-executing task from observing updates in
the page buffer from a previous execution attempt. Clearing \texttt{\underline{shadow}}
entries as part of the second phase commit (Line 13), ensures that all accesses to
protected variables in subsequent tasks correctly access their original memory
locations in \texttt{\underline{store}}.
%
%The non-volatile commit bit persists
%across power failures and \sys repeatedly attempts to commit until it succeeds,
%despite arbitrary power failures.
\subsection{Dynamic vs. Static Privatization}
\TODO{Put paragraph about Alpaca's static privatization algo vs. Coala's dynamic privatization algo}
\TODO{pass by ref might not work the right way}
\begin{verbatim}

>>>>

Paging moves data between non-volatile and volatile memory. A page has a home location in non-volatile memory which we refer to as the page's \texttt{\underline{store}} (underlined locations are non-volatile).
As a task accesses pages, they are swapped from their home location
into the fully\hyp{}associative \texttt{working} page buffer in volatile memory. The \texttt{working} buffer is allocated to take up the volatile memory that remains after space for stack is allocated.
%
\sys uses a two-phase commit to write dirty pages from the \texttt{working} buffer back to their home location in \texttt{\underline{store}} atomically, ensuring data consistency even with power failures.
%
When any dirty page is swapped out from the \texttt{working} buffer due to eviction or task transition from last coalesced task, the first phase writes the page to a direct\hyp{}mapped non-volatile \texttt{\underline{shadow}} buffer, and the second phase writes it from \texttt{\underline{shadow}} to \texttt{\underline{store}}.
%
Unlike prior work~\cite{chain,alpaca} that moves scattered variables one at a time between volatile and non-volatile memory, \sys's page-based design moves contiguous blocks of data.
%
\sys takes full advantage of hardware\hyp{}accelerated block copies using Direct Memory Access (DMA) hardware in common microcontrollers --- e.g. 5x fewer cycles on MSP430FRxxxx.

@@

A \sys task must access protected non-volatile variables through a restricted memory interface, as will be described in Section~\ref{sec:coala_api}. The interface includes \texttt{RP(var)}, to read the value of variable {\tt var}, and \texttt{WP(var)} to assign a value to variable {\tt var}. The implementation of {\tt RP} is shown in Algorithm~\ref{algo:rwar}, and {\tt WP}'s implementation is similar except for marking the accessed page as dirty.
%
{\tt RP} and {\tt WP} operations translate a variable's physical address in non-volatile memory into a \emph{virtual address} in the page buffer in volatile memory. In \sys, a virtual address is composed of a \emph{page tag} that identifies the page and a \emph{page offset} that identifies a byte.

@@

\begin{algorithm}[t]
	\caption{\texttt{RP}(variable $v$)}
	\label{algo:rwar}
	\scriptsize
	%\small
	\begin{algorithmic}[1]
		\State $t \leftarrow \Call{GetTag}{v}$ 
        \State $i \gets \{ j\ |\ \Call{GetTag}{\texttt{working}[j]} = t \}$ \Comment{Look in the page buffer}
		\If { $i = \emptyset$ } \Comment{Is the variable in a resident page?}
		\State	$i \gets \Call{PageFault}{t}$ \Comment{Page in from non-volatile memory}
		\EndIf
		\State $o \leftarrow \Call{GetOffset}{v}$ 		
		\State \Return \texttt{working}[$i$][$o$]  \Comment{Return directly from page buffer}
	\end{algorithmic}
\end{algorithm}

After address translation, \sys attempts to access the protected variable's location in the volatile paging buffer. \sys keeps track of the page tags for the pages currently resident in the paging buffer. When a task accesses a variable, it compares the variable's page tag to tags of the pages in {\tt working} (Line 2).
%
%If the accessed variable's page is resident in the page buffer,
%then the tag of one of the pages in this buffer is equal to the accessed
%variable's page tag.
%
If the accessed variable's page tag is not found in the page buffer, the operation incurs a {\em page fault} (Line 4). At a page fault, \sys must swap out one of the pages that is resident in the page buffer (to the non-volatile \texttt{\underline{shadow}}) if \texttt{working} is full, and swap in the variable's page.
%
The byte is accessed in the page buffer at the index of the resident page and the variable's page offset (Lines 5-6).

@@

\begin{algorithm}
	\caption{\texttt{PageFault}(tag $t$)}
	\label{algo:pagefault}
	\scriptsize
	%\small
	\begin{algorithmic}[1]
        \State \LeftComment{$i_\text{victim}$: volatile index to next free slot or page to be evicted, set to 0 at boot}
        \State \LeftComment{\texttt{\underline{shadowCount}}: non-volatile, reset to 0 on boot unless in phase 2 commit}
        \State $t_\text{victim} \gets \Call{Tag}{\texttt{working}[i_\text{victim}]}$
		\If {\Call{isDirty}{working[$i_\text{victim}$]} }	\Comment{Are we evicting a dirty page?}
		\State $\texttt{\underline{shadow}}[t_\text{victim}] \xleftarrow{\text{DMA}} \texttt{working}[i_\text{victim}]$
            \Comment{Swap out the page}
        \If {$t_\text{victim} \not\in \texttt{\underline{shadowList}}$} \Comment{Keep track as list for quick traversal}
            \State $\texttt{\underline{shadowList}}[\texttt{shadowCount++}] \gets t_\text{victim}$
        \EndIf
		\EndIf
		\If {\Call{exists}{\underline{shadow}[$t$]}} \Comment{Had the page been swapped out earlier?}
		\State $\texttt{working}[i_\text{victim}] \xleftarrow{\text{DMA}} \texttt{\underline{shadow}}[t]$
            \Comment{Get page from commit buffer}
		\Else
		\State $\texttt{working}[i_\text{victim}] \xleftarrow{\text{DMA}} \texttt{\underline{store}}[t]$
            \Comment{Get page from backing store}
		\EndIf 
        \State $i_\text{victim} \gets (i_\text{victim} + 1) \mod |\texttt{pageBufLength}|$ \Comment{FIFO eviction policy}
	\end{algorithmic}
\end{algorithm}

If the \texttt{working} buffer is full, a page fault on a memory access requires \sys to swap out one of the pages in the volatile buffer, the \emph{victim page}, preserving updates made to that page, and to copy the accessed page of protected data into the page buffer. Algorithm~\ref{algo:pagefault} shows how \sys handles a page fault. A victim page in the \texttt{working} buffer is selected for eviction according to \emph{first-in-first-out} replacement policy (Lines 2 and 12). If the page being swapped out is not dirty and the page being accessed had not been dirtied and swapped out to the \texttt{shadow} buffer since last boot, then handling a page fault consists of copying the accessed page from \texttt{\underline{store}} into the \texttt{working} buffer (Line 11). Otherwise, the page fault handler must take the following extra steps.
%
If a task modified any byte in the victim page (i.e., using \texttt{WP}), then the page is dirty. Before swapping in the accessed page, \sys must first preserve the updated values in the dirty page so that they can be committed back to their locations in memory when the dynamic task ends (but not earlier). To preserve dirty values, \sys copies the victim page to the non-volatile \texttt{\underline{shadow}} buffer (Line 5) and records the page's tag in a non-volatile list (Lines 6-7) to support iteration over shadow pages having traversing all slots in the buffer, necessary in the second phase of the commit.
%
%Dirty pages remain in {\tt \underline{shadow}} buffer until the
%dynamic task completes, at which point \sys commits the modified
%pages from {\tt \underline{shadow}} back to \texttt{\underline{store}}.
%
If the accessed page was previously modified and swapped out since the last boot (Line 8), the most recent version of the page is in {\tt
\underline{shadow}}, not \texttt{\underline{store}} (Line 9).

@@

When the last task in a coalesced group completes, \sys must commit \emph{all} dirty pages in the \texttt{working} buffer and the swapped out pages in \texttt{\underline{shadow}} buffer back to their original locations in \texttt{\underline{store}}. If a task is allowed to (re-)execute after only \emph{some} pages have been committed to \texttt{\underline{store}}, that task might see protected variables in an inconsistent state.

To make the commit atomic, \sys commits the pages in two phases, implemented by Algorithm~\ref{algo:commit}.
%
The first phase copies dirty pages from the \texttt{working} buffer to the non-volatile \texttt{\underline{shadow}} buffer (Line 4). The second phase commits pages from \texttt{\underline{shadow}} to \texttt{\underline{store}} (Line 12).
If power fails during the first phase, the whole commit is aborted, and the dynamic coalesced task restarts from its first static task. If power fails in the second phase, the commit will be safely resumed on next boot.
The non-volatile metadata for the second phase are: the \texttt{\underline{committing}} bit, which is set before the first page is committed (Line 9) and cleared after the last page is committed (Line 16); \texttt{\underline{shadowCount}} counter, which is cleared once the phase completes (Line 14); \texttt{\underline{commitIndex}}, which indexes the next page to be committed (Line 11) and is cleared at the end of the phase (Line 15).

\sys non-volatile \texttt{\underline{shadowList/shadowCount/commitIndex}} implementation is a more efficient alternative to a simpler implementation that keeps a valid bit in each entry of \texttt{\underline{shadow}} and, on every boot into second phase, traverses all entries, committing the pages with a set bit, and clearing the bit. During the second commit phase, pages are not physically copied from \texttt{\underline{shadow}} to \texttt{\underline{store}}, but a table pointing at the right buffer for each page is updated. This is explained in more details in Section~\ref{sec:implementation}.

\begin{algorithm}[t]
	\caption{Two-phase commit}
	\label{algo:commit}
	\scriptsize
	%\small
	\begin{algorithmic}[1]
        \Procedure{CommitPhase1}{} \Comment{Invoked from \texttt{next\_task} on last coalesced task}
            \For {$i \in 0..|\texttt{working}|-1$}
                \State $t \gets \Call{Tag}{\texttt{working}[i]}$
                \State $\texttt{\underline{shadow}}[\Call{Tag}{\texttt{working}[i]}] \xleftarrow{\text{DMA}} \texttt{working}[i]$
                \State $\texttt{\underline{shadowList}}[\texttt{\underline{shadowCount}}] \gets t$
                \State $\texttt{\underline{shadowCount}} \gets \texttt{\underline{shadowCount}} + 1$
            \EndFor
            \State \Call{CommitPhase2}{}
        \EndProcedure
        \Procedure{CommitPhase2}{}
            \State $\texttt{\underline{committing}} \gets \textbf{true}$
            \While{\texttt{\underline{commitIndex}} < \texttt{\underline{shadowCount}}} \Comment{\texttt{\underline{commitIndex}} set to 0 at first boot}
                \State $t \gets \texttt{\underline{shadowList}}[\texttt{\underline{commitIndex}}]$
                \State \Call{commitToStore}{\texttt{\underline{shadow}}[t]}
                \State $\texttt{\underline{commitIndex}} \gets \texttt{\underline{commitIndex}} + 1$
            \EndWhile
            \State $\texttt{\underline{shadowCount}} \gets 0$
            \State $\texttt{\underline{commitIndex}} \gets 0$
            \State $\texttt{\underline{committing}} \gets \textbf{false}$
        \EndProcedure
        \Procedure{OnBoot}{} \Comment{Invoked on every boot, before transitioning to a task}
            \If { \texttt{\underline{committing}} } \Call{CommitPhase2}{}
            \EndIf 
            \State $\texttt{\underline{shadowCount}} \gets 0$, $i_\text{victim} \gets 0$
        \EndProcedure
	\end{algorithmic}
\end{algorithm}

@@

\textbf{Memory Consistency.} \sys's paging mechanism ensures that a task only ever executes using consistent protected data. During task execution, modifications to protected data do not affect their original memory locations, because a task reads and writes the volatile \texttt{working} buffer only, and modified pages are kept in \texttt{\underline{shadow}} until commit.
A power failure erases the contents of the page buffer, preventing a re-executing task from observing updates in the page buffer from a previous execution attempt.
Clearing \texttt{\underline{shadowCount}} as part of the second phase commit (Line 14), ensures that all accesses to protected variables in subsequent tasks correctly access their original memory locations in \texttt{\underline{store}}.
\sys assumes that the write to a non-volatile word in memory is atomic.
%
%The non-volatile commit bit persists
%across power failures and \sys repeatedly attempts to commit until it succeeds,
%despite arbitrary power failures.
\subsection{Dynamic vs. Static Privatization}
\TODO{Put paragraph about Alpaca's static privatization algo vs. Coala's dynamic privatization algo}
\TODO{pass by ref might not work the right way}
\begin{verbatim}
